---
title: "SSCM Exercise 5"
author: "Nikolaus Czernin - 11721138"
output: pdf_document
fig_height: 4 

---

```{r}
library("tidyverse")
library(knitr)

# Custom print function
print_ <- function(...) print(paste(...))

set.seed(11721138)
```
# Task 1
```{r}
x1 <- c(-0.673, -0.584, 0.572, -0.341, -0.218, 0.603, -0.415, -0.013, 0.763, 0.804, 0.054, 1.746, -0.472, 1.638, -0.578, 0.947, -0.329, -0.188, 0.794, 0.894, -1.227, 1.059)
x2 <- c(0.913, -0.639, 2.99, -5.004, 3.118, 0.1, 1.128, 0.579, 0.32, -0.488, -0.994, -0.212, 0.413, 1.401, 0.007, 0.568, -0.005, 0.696)

data.frame(
  Sample=c("x1", "x2"), 
  Mean = c(mean(x1), mean(x2))
) %>% kable()

```

## Plotting
```{r}

# Means
mean_x1 <- mean(x1)
mean_x2 <- mean(x2)

boxplot(x1, x2, 
        horizontal = TRUE, 
        names = c("x1", "x2"),
        main = "Boxplot with means as dots",
        col = c("lightblue", "lightcoral")
        )

points(mean_x1, 1, cex=2 , pch = 18)
points(mean_x2, 2 , cex=2, pch = 18)


```
Looking at the boxplots, the means seem to be pretty much the same almost. 
The first sample has fewer outliers though, a wider outer quantile distance and is more 
right-skewed. 

## Bootstrapping
```{r BS non-param}

bootstrap.t.indep <- function(data.1, data.2, m=10000, ...) {
  # compute the original test-statistic
  original.t <- t.test(x1, x2)$statistic
  # generate bootstrap samples
  bs.samples.1 <- lapply(1:m, function(x) sample(data.1, replace = TRUE))
  bs.samples.2 <- lapply(1:m, function(x) sample(data.2, replace = TRUE))
  # compute the estimated parameters of the samples
  bs.ts <- sapply(1:m, function(i) t.test(bs.samples.1[[i]], bs.samples.2[[i]])$statistic)
  # we now have a student's t distribution
  # we can extract the p-value from it
  p.value = mean(abs(bs.ts) >= abs(original.t))
  # also get the confidence intervals, i.e. the quantiles of the t-stats
  ci.95 <- quantile(bs.ts, c(0.05, 0.95))
  ci.99 <- quantile(bs.ts, c(0.01, 0.99))
  # return whatever was created here
  list(p.value = p.value, ci.95 = ci.95, ci.99 = ci.99)
}

bootstrap.t.indep(x1, x2, m=30)
```
Here I define a function for non-parametric bootstrapping as a test. 
The function first computes the estimator, in this case the student's t-test on 
the original two samples. It then iterates m times and each time applys
non-parametric bootstrapping by sampling from the original samples with replacement 
to create 2 new samples, which are compared again with t tests. 
The t statistics of these m tests form a sort of student's t distribution, 
for wich I calculate a p value and confidence intervals. 

```{r BS centered}
bootstrap.t.centered <- function(data.1, data.2, m=10000) {
  # compute the mean of the combined samples
  mean.center <- mean(c(data.1, data.2))
  print("Centering the means by" %>% paste(-mean.center %>% round(3)))
  # center the 2 samples
  data.1 <- data.1 - mean.center
  data.2 <- data.2 - mean.center
  # apply the independent bootstrap esimation
  bootstrap.t.indep(data.1, data.2, m=m)

  }

bootstrap.t.centered(x1, x2, m=30)

```
The function here works similarly to the one before, it only applies 
combined mean centering before generating samples and applying the estimators beforehand. 

## Applying the functions on t-tests
```{r}

results.bootstrap.t.indep <- bootstrap.t.indep(x1, x2)
results.bootstrap.t.centered <- bootstrap.t.centered(x1, x2)

res <- data.frame(
  Independed.Bootstrap.Sampling=c(
    results.bootstrap.t.indep$ci.95 %>% round(4) %>% paste(collapse = " - "), 
    results.bootstrap.t.indep$ci.99 %>% round(4) %>% paste(collapse = " - "),
    results.bootstrap.t.indep$p.value
  ),
  Mean.Centered.Bootstrap.Sampling=c(
    results.bootstrap.t.centered$ci.95 %>% round(4) %>% paste(collapse = " - "), 
    results.bootstrap.t.centered$ci.99 %>% round(4) %>% paste(collapse = " - "),
    results.bootstrap.t.centered$p.value
  )
) 

rownames(res) <- c("95% Conf.Level", "99% Conf.Level", "p-Value")
res %>% 
  kable(caption="Comparison of Bootstrap Sampling methods on t-tests")
  
  
  

```
Im getting p-values well over 0.9, so we are far from being able to reject the 
null-hypothesis of the samples having different means.  


## Permutation

```{r}
permutation_test <- function(data.1, data.2, m=4){
  # get the t statistic of the original samples
  t.original <- t.test(data.1, data.2)$statistic
  # combine the datasets
  data.combined <- c(data.1, data.2)
  # create bootstrap samples from the combined data
  t.bs <- sapply(1:m, function(x) {
    # create a permutation of the combined dataset
    permutation_idx <- sample(1:length(data.combined), size = length(data.1), replace = FALSE)
    # get 2 samples at the sizes of the original samples
    permutation.1 <- data.combined[permutation_idx]
    permutation.2 <- data.combined[-permutation_idx]
    # perform a t test and return the t statistic
    t.test(permutation.1, permutation.2)$statistic
  })
  
  # calculate the p value
  p.value <- mean(abs(t.bs) >= abs(t.original))
  # also get the confidence intervals, i.e. the quantiles of the t-stats
  ci.95 <- quantile(t.bs, c(0.05, 0.95))
  ci.99 <- quantile(t.bs, c(0.01, 0.99))
  # return whatever was created here
  list(p.value = p.value, ci.95 = ci.95, ci.99 = ci.99)
  
}



results.permutation_test <- permutation_test(x1, x2, m=10000)

res <- data.frame(
  "Independed Bootstrap Sampling"=c(
    results.permutation_test$ci.95 %>% round(4) %>% paste(collapse = " - "), 
    results.permutation_test$ci.99 %>% round(4) %>% paste(collapse = " - "),
    results.permutation_test$p.value
  )
) 

rownames(res) <- c("95% Conf.Level", "99% Conf.Level", "p-Value")
res %>% 
  kable(caption="Permutation Sampling method on t-tests")
  
  
```
Instead of adapting the first test function I made, I implemented a new function. 
It again creates a function that iterates m times and applies a t-test on random 
index permutations of the combined dataset. Using the same formula as before I 
calculate the p values and the confidence intervals on the t statistics of the 
m students t test statistics.  

The p value of the test is again ~0.9, so we are not rejecting the Null-hypothesis 
of the two samples having different means. 

## T-Test and Wilcoxon test
```{r}
t.test(x1, x2)
wilcox.test(x1, x2)
```
The results of the base R `t.test()` also has a p-value of ~0.9, which is in line 
with my simulated tests and my conclusion of keeping the null-hypothesis. 

The p-value of the Mann-Whitney test is ~0.66, which is also high enough for us 
not to be able to reject the null-hypothesis. 




# Task 2

```{r}
n <- 200

# define function to sample the used variables from their distributions
x1 <- rnorm(n, mean = 2, sd = sqrt(3)) 
x2 <- runif(n, 2, 4)                   
x3 <- runif(n, -2, 2)                  
e <- rt(n, df = 5)                     

# define the function y
# Model: y = 3 + 2*x1 + x2 + e
y <- 3 + 2 * x1 + x2 + e               

# Generate a sample of size n
data <- data.frame(y = y, x1 = x1, x2 = x2, x3 = x3)

data %>% head()


```
I define functions that source the variables from their respective distributions 
and can be passed to the y function to create the model output. 
I use piping to create a dataframe of size 200 with all parameters and the output 
variable y. 


## Residual bootstrap

```{r}
# custom rmse function
get_rmse <- function(model, r=4) sqrt(mean(model$residuals^2)) %>% round(r)

# fitting the linear model
lin <- lm(y ~ x1 + x2 + x3, data=data)
lin %>% summary()

yhat <- predict(lin, data %>% select(-y))
plot(
  data$y,
  yhat,
  main="Prediction plot of the linear model — RMSE:" %>% paste(., get_rmse(lin)), 
  ylab="predicted", xlab="observed"
)
abline(coef = c(0,1), col="red")
```
In the linear model, all included variables are significant, even x3, which played 
no part in creating the output variable and essentially distorts the data, but its 
coefficient is also the smallest.  
Just visually, the fit is fine. I also report the RMSE in the heading.

## Residual Bootstrap Regression
```{r Residual Bootstrap Regression}
# get the residuals
residuals <- lin$residuals
# create the bootstrap sample

residual.bootstrap.coeffs <- function(data, yhat, residuals, m=1000){
    # generate new residuals via bootstrapping, m times
    residuals.resampled <- sapply(1:m, function(j) sample(residuals, replace = TRUE))
    # we now get a matrix (n, m), where n is the size of the original sample
    # and m is the number of bootstraps were doing, e.g. 1000
    # create new predictions by adding the sampled residuals to the yhat
    # do this by adding the columns of residuals.resampled to the yhat
    yhat.bootstrapped <- apply(residuals.resampled, 2, function(rs) rs + yhat)
    # now we have another (n, m) matrix
    # fit a new model to each of the bootstrapped predictions, 
    # without using e (epsilon)
    new.models <- lapply(1:m, function(j) lm(yhat.bootstrapped[,j] ~ x1 + x2 + x3, data=data))
    # we now have m different model objects and now we extract their coefficients
    # return the coefficients of the bootstrapped model
    coeffs <- lapply(new.models, coef) %>% 
      do.call(rbind, .)
    # we now have a matrix of size (m, 4), where the 4 are the coefficients incl intercept
    coeffs
    # now extract the quantiles for each of the 4 coefficients
    apply(coeffs, 2, function(b) quantile(b, probs=c(0.05, 0.95)))
    

}

residual.bootstrap.coeffs(data, yhat, residuals)

```
In the chunk above I define a test that takes the residuals of a linear model 
and creates m bootstrap samples by sampling with replacement from them. 
These resampled residuals are then added as not-so-random noise to the 
predicted response of the former linear model. By then trying to again fit 
m different models to these adjusted y values, we are simulating a correction of 
the models to the y values corrected by the residuals, i.e. nudging them around in 
the range of the actually observed values.  
The m different models then have each 4 coefficients, and each of them has m different 
versions, for which I then grab the confidence intervals.  

Since the confidence interval of the x3 variable is centered around 0, I presume 
that it is not really necessary  for the model, even though it was found as 
significant by the `lm()` function. 



## Pairs Bootstrap Regression
```{r Pairs Bootstrap Regression}

# create the bootstrap sample
pairs.bootstrap.coeffs <- function(data, m=1000){
    bootstrap.samples <- lapply(1:m, function(j) data[sample(1:nrow(data), replace = TRUE),])
    # print(bootstrap.samples %>% length())
    # print(bootstrap.samples[[2]])
    # we now got m different resampled dataframes
    # now refit new models with them
    new.models <- lapply(bootstrap.samples, function(d) lm(y ~ x1 + x2 + x3, data=d))
    # we now have m models
    # now extract their coefficients and turn them into a matrix
    coeffs <- lapply(new.models, coef) %>% 
      do.call(rbind, .)
    # now extract the quantiles for each of the 4 coefficients
    apply(coeffs, 2, function(b) quantile(b, probs=c(0.05, 0.95)))
    

}

pairs.bootstrap.coeffs(data)

```

In this chunk I define a test that resamples from the original dataset, instead of the 
residuals. It samples rows of the dataset with replacement, exactly as many rows 
as the original dataset had. Then it fits the new data in a new linear model. 
All that is repeated m times and then coefficients' distributions' confidence intervals 
are returned.  
In this case, the confidence interval of the variable x3 is no longer surrounding 0, 
so if I didnt know any better, I'd presume that this time around that the 
true model likely even includes this variable. Other than that the confidence intervals 
are very close to what they have been in the residual approach. 


## Approach comparison
The first approach sampled from the results of an initial model fitting, whereas 
the second sampled the original data. I there assume, that the first approach 
makes some kind of assumption that the model was already doing something right, 
like for example variable selection. The second approach on the other hand has 
never even seen any model before.  
The first approach, while bootstrap sampling always uses the same predictors and 
input data, but always to predict something slightly different, whereas the 
second always tries to predict the same output using different permutations of 
its original input everytime. 



# Summary

Bootstrapping is a method where we create many new samples from our data by either resampling the data directly (non-parametric) or generating samples based on an assumed model (parametric). We use these samples to calculte confidence intervals, test statistics, or model estimtes. Parametric bootstrap assumes a specific distribution, which can give more precise results if the assumption is correct but may give misleading results if the model is wrong. Non-parametric bootstrap is flexible and works well for any dataset, but it may require more computational effort and does not smooth out sampling noise. Bootstrapping is powerful because it relies on the data itself to understand variability and does not need complex formulas for confidence bounds or tests.








