---
title: "SSCM Exercise 2"
author: "Nikolaus Czernin - 11721138"
output: pdf_document
fig_height: 4 

---

```{r}
library("tidyverse")
# install.packages("FRACTION")
library("FRACTION")
```


# Linear Congruential Random Number Generator

```{r}
# prepare the PRNG function
lcrng <- function(n, m, a, c=0, x0){
  us <- numeric(n)
  # keep an extra vector for the x values
  xs <- numeric(n)
  for (i in 1:n){
    x0 <- (a * x0 + c) %% m
    xs[i] <- x0
    us[i] <- x0 / m
  }
  list("u"=us, "x"=xs)
}


visualize_random_numbers <- function(n, m, a, c, x0, title=""){
  # generate random numbers with given params
  prns <- lcrng(n, m, a, c, x0)
  df <- data.frame(
    i = 1:n,
    u = prns$u,
    x = prns$x,
    n = n,
    m = m,
    a = a,
    c = c,
    x0 = x0
  ) 
  
  # prepare a double-plot window
  # set up the layout with 2 rows and 2 columns
  layout(matrix(c(1, 2, 3, 3, 4, 4), nrow = 3, byrow = TRUE))
  
  par(mar = c(3, 3.5, 2.5, 2),
      mgp = c(1.5, 0.5, 0))   
  
  # create the plots 
  df$u %>% hist(main="")
  plot(df$i, df$u, xlab="", ylab="Random Number", ylim=c(0, 1))
  
  # for the line plot, only use the first 20 numbers
  df_head <- df %>% head(20)
  plot(df_head$i, df_head$u, type="b", xlab="", ylab="Random Numbers", ylim=c(0, 1))
  first_cycle <- filter(df, x==x0) %>% head(1) %>% .$i
  abline(v=first_cycle, col="blue")

  # plot also the x-values
  plot(df_head$i, df_head$x, type="b", xlab="", ylab="xx-values")
  abline(v=first_cycle, col="blue")
  
  # make a custom title with the parameters
  text <- paste0(title, ":   m=", m, "   a=", a, "   c=", c, "   x0=", x0, "\n")
  mtext(text, side=3, outer=TRUE, line=-3)

}

# visualize_random_numbers(200, 9, 8, 7, 6, "Plot A")

```
In this chunk, I defined a PRNG function that uses the Linear Congruential Random Number Generator algorithm. 
It expects all parameters as arguments. 
I defined another function that visualizes the result using a histogram, a scatterplot
and a lineplot of the first values to visualize a sequence-loop. It also visualizes the x-values for better analysis.

```{r}
visualize_random_numbers(200,  64,   8,  18,  13, "Plot A")
visualize_random_numbers(200,  27,   3,  18,  13, "Plot B")
visualize_random_numbers(200,  64,  56,  18,  13, "Plot C")
visualize_random_numbers(200,  25,  15,  18,  13, "Plot D")
```
Plots A through D above all loop very early. What they have in common is that m and a have common denominators. 
After numbers in the sequence, the values stagnate.
  
The blue lines mark the first iteration, where the generated x-value is equal to the initial $x_0$ value, 
which is the latest point of a cycle restarting, though a cycle may start even earlier too.


```{r}
visualize_random_numbers(200,  65,   8,  18,  13, "Plot E")
visualize_random_numbers(200,  28,   3,  18,  13, "Plot F")
visualize_random_numbers(200,  61,  56,  18,  13, "Plot G")
visualize_random_numbers(200,  26,  15,  18,  13, "Plot H")

```
When using m values that have no common denominator with a the loops are created later, as seen in plots E through H. 



```{r}
visualize_random_numbers(200,  67,   8,  17,  13, "Plot I")
visualize_random_numbers(200,  11,   3,  17,  13, "Plot J")
visualize_random_numbers(200,  3119,  56,  17,  13, "Plot K")
visualize_random_numbers(200,  227,  15,  17,  13, "Plot L")


```
As seen in Plots I through L, when using large prime numbers as m, i.e. 3-digit value or higher, 
there are no apparent cycles in the random number sequences anymore.

```{r}
visualize_random_numbers(200,  1,  56,  1,  1, "Plot M")
visualize_random_numbers(200,  3119,  56,  0,  0, "Plot N")
visualize_random_numbers(200,  3119,  56,  1,  0, "Plot O")
visualize_random_numbers(200,  3119,  56,  0,  1, "Plot P")

```
Plot M shows that if m is 1, the random numbers will be all zeros, as modulus of 1 will always be 0 on integers.
Plot N shows that having both c and $x_0$ equal to zero leads to a sequence producing only zeros. 
If either are unequal to zero, you get a random sequence with seemingly no cycles, as seen in plots O and P.


```{r}
visualize_random_numbers(200,  1,  56,  17,  13, "Plot K")
```

