---
title: "SSCM Exercise 4"
author: "Nikolaus Czernin - 11721138"
output: pdf_document
fig_height: 4 

---

```{r}
library("tidyverse")
library(knitr)

# Custom print function
print_ <- function(...) print(paste(...))

set.seed(11721138)
```

```{r}
# Define our starting sample X
X <- "4.94 5.06 4.53 5.07 4.99 5.16 4.38 4.43 4.93 4.72 4.92 4.96" %>% strsplit(" ") %>% unlist() %>% as.numeric() %>% print()
N <- length(X)

```
Being lazy, I used strsplit() to get the values from the given sample text.  

```{r}
ecdf <- function(n) c(0, seq(1,n)/n)
```

# Task 1

## Number of bootstrap samples
If we had a sample $X={1, 2, 3}$, with N=3, we would generate a bootstrap sample by sampling from X with replacement. The first number in the new sample would have thus N=3 possible values. For each value of the first sample, each other value could have also 1 out of 3 values. The number of different samples is therefore $3*3*3=N^N$, which gives us the formula for any size of N (skipping the formal proof here).  
The number of possible bootstrap samples is therefore $N^N=12^12=$`r 12^12`.

## Mean and median
```{r}
X.mean <- mean(X)
X.mean %>% print_("Mean of X:", .)
X.median <- median(X)
X.median %>% print_("Median of X:", .)
```

## Generate 2000 bootstrap samples
```{r}
m <- 2000
bootstrap.samples <- lapply(1:m, function(x) sample(X, replace = TRUE))
```

### Analyse their means
```{r}
bootstrap.means <- sapply(bootstrap.samples, function(s) mean(s))
```
I use lapply to loop 2000 times and create a new sample with replacement every time and then save the means of the sample using sapply instead.

```{r}
mean20 <- mean(bootstrap.means[1:20]) %>% print_("Mean of the first 20 sample means:", .)
mean200 <- mean(bootstrap.means[1:200]) %>% print_("Mean of the first 200 sample means:", .)
mean2000 <- mean(bootstrap.means[1:2000]) %>% print_("Mean of the first 2000 sample means:", .)

```


```{r}
hist(bootstrap.means[1:20], breaks = seq(min(X), max(X), 0.05), main="Histogram of 2000 bootstrap means")
abline(v=X.mean, col="red")
hist(bootstrap.means[1:200], breaks = seq(min(X), max(X), 0.01), main="Histogram of 2000 bootstrap means")
abline(v=X.mean, col="red")
hist(bootstrap.means[1:2000], breaks = seq(min(X), max(X), 0.005), main="Histogram of 2000 bootstrap means")
abline(v=X.mean, col="red")



density(bootstrap.means) %>% plot(main="Smooth histogram of 2000 bootstrap means")
abline(v=X.mean, col="red")

```
At increasing sample size m we cna clearly observe the distribution of the means steadily getting the shape of a normal distirbution, outlining the central limit theorem.  
I also plotted the curve using the density() of the means and plotting it as a line. 

```{r Quantiles}

data.frame(
  k = c(20, 200, 2000, "True confidence interva"),
  Q025 = c(
    quantile(bootstrap.means[1:20], 0.025),
    quantile(bootstrap.means[1:200], 0.025),
    quantile(bootstrap.means[1:2000], 0.025),
    t.test(X)$conf.int[1]
    ),
  Q975 = c(
    quantile(bootstrap.means[1:20], 0.975),
    quantile(bootstrap.means[1:200], 0.975),
    quantile(bootstrap.means[1:2000], 0.975),
    t.test(X)$conf.int[2]
    )
)

```
By chance, the quantiles of the mean of the first 200 means appear to be an even close match to the "true" distributions, or rather a normal distribution with the same mean and standard deviation as the original sample, than the full 2000 bootstrap samples. 


### Analyse their medians
```{r}
bootstrap.medians <- sapply(bootstrap.samples, function(s) median(s))
```

```{r}
mean.median20 <- mean(bootstrap.medians[1:20]) %>% print_("Mean of the first 20 sample medians:", .)
mean.median200 <- mean(bootstrap.medians[1:200]) %>% print_("Mean of the first 200 sample medians:", .)
mean.median2000 <- mean(bootstrap.medians[1:2000]) %>% print_("Mean of the first 2000 sample medians:", .)

```


```{r}
hist(bootstrap.medians[1:20], breaks = seq(min(X), max(X), 0.05), main="Histogram of 2000 bootstrap medians")
abline(v=X.median, col="red")
hist(bootstrap.medians[1:200], breaks = seq(min(X), max(X), 0.01), main="Histogram of 2000 bootstrap medians")
abline(v=X.median, col="red")
hist(bootstrap.medians[1:2000], breaks = seq(min(X), max(X), 0.005), main="Histogram of 2000 bootstrap medians")
abline(v=X.median, col="red")


density(bootstrap.medians) %>% plot(main="Smooth histogram of 2000 bootstrap medians")
abline(v=X.mean, col="red")

```
Regarding the medians, the Central Limit Theorem does not show at all, regarding the overall mean of medians. With more samples I feel like there might be a pattern of multiple little curves in the shape of normal distirbutions for each value in the original sample.  
I also plotted the curve again using the density() of the medians and plotting it as a line. 

```{r Quantiles}

data.frame(
  k = c(20, 200, 2000, "True confidence interval"),
  Q025 = c(
    quantile(bootstrap.medians[1:20], 0.025),
    quantile(bootstrap.medians[1:200], 0.025),
    quantile(bootstrap.medians[1:2000], 0.025),
    t.test(X)$conf.int[1]
    ),
  Q975 = c(
    quantile(bootstrap.medians[1:20], 0.975),
    quantile(bootstrap.medians[1:200], 0.975),
    quantile(bootstrap.medians[1:2000], 0.975),
    t.test(X)$conf.int[2]
    )
)

# t.test(X)$conf.int[1]
```
The mean medians' quantiles dont match the confidence interval of the original distribution as closely as the mean means did. 





# Task 2
## Generate sample
```{r}
set.seed(1234)

x.clean <- rnorm(1960)
x.cont <- runif(40, min=4, max=5)
# total data
x <- c(x.clean,x.cont)

set.seed(11721138)

```

## Estimate mean and median
```{r}
alpha = 0.05

x.median <- median(x)
x.mean <- mean(x)
x.mean.trimmed <- mean(x, trim = 0.05)  

x.clean.median <- median(x)
x.clean.mean <- mean(x)
x.clean.mean.trimmed <- mean(x, trim = 0.05)

```


```{r}
m <- 2000
```

```{r}
bootstrap.nonparam <- function(data, estimator, m=2000, ...) {
  bootstrap.samples <- lapply(1:m, function(x) sample(data, replace = TRUE))
  stats <- sapply(bootstrap.samples, function(y) estimator(y, ...))
  # Standard error
  se <- sd(stats)
  # 95 percentile confidence interval
  ci <- quantile(stats, probs = c(0.025, 0.975))
  list(se = se, ci = ci)
}

x.bs.median <-  bootstrap.nonparam(x, median)
x.bs.mean <-  bootstrap.nonparam(x, mean)
x.bs.meant <-  bootstrap.nonparam(x, mean, trim=0.05)

x.clean.bs.median <-  bootstrap.nonparam(x.clean, median)
x.clean.bs.mean <-  bootstrap.nonparam(x.clean, mean)
x.clean.bs.meant <-  bootstrap.nonparam(x.clean, mean, trim=0.05)

data.frame(
  Sample = c(rep("x", 3), rep("x.clean", 3)),
  Estimator = c("Median", "Mean", "Mean Trimmed") %>% rep(2),
  Standard.Error = c(x.bs.median$se, x.bs.mean$se, x.bs.meant$se, x.clean.bs.median$se, x.clean.bs.mean$se, x.clean.bs.meant$se) %>% sapply(round, 4),
  Confidence.Intervals=list(x.bs.median$ci, x.bs.mean$ci, x.bs.meant$ci, x.clean.bs.median$ci, x.clean.bs.mean$ci, x.clean.bs.meant$ci) %>% sapply(function(c) paste(c[1] %>% round(4), ", ", c[2] %>% round(4)))
) %>% 
  print() %>% 
  kable()

```
In my `bootstrap.nonparam()` function I  use lapply to sample from the originally given data m times and apply the given estimator, i.e. `median`, `mean` or `mean` with an extra parameter. 
I then compute the standard error and the 95% quantiles.  




## Parametric bootstrap
```{r}
# Function to perform parametric bootstrap and calculate statistics
parametric_bootstrap <- function(data, estimator, center, scale, m=2000, ...) {
  # generate samples 
  bootstrap.samples <- lapply(1:m, function(x) rnorm(m, mean = center, sd = scale))
  # get stats (mean, median, ...)
  stats.estimated <- sapply(bootstrap.samples, function(y) estimator(y, ...))
  # get the stat of the original sample
  stat.original = estimator(data, ...)
  # get bias, standard error, and 95 percentile confidence interval
  bias <- mean(stats.estimated) - stat.original
  se <- sd(stats.estimated)
  ci <- quantile(stats.estimated, probs = c(0.025, 0.975))
  bias_corrected_estimate <- stat.original - bias

  list(bias = bias, se = se, ci = ci, bias_corrected_estimate = bias_corrected_estimate)

}


x.center <- mean(x)
x.scale <- sd(x)
x.clean.center <- mean(x.clean)
x.clean.scale <- sd(x.clean)


x.bsp.median <-  parametric_bootstrap(x, median, x.center, x.scale)
x.bsp.mean <-  parametric_bootstrap(x, mean, x.center, x.scale)
x.bsp.meant <-  parametric_bootstrap(x, mean, x.center, x.scale, trim=0.05)

x.clean.bsp.median <-  parametric_bootstrap(x.clean, median, x.clean.center, x.clean.scale)
x.clean.bsp.mean <-  parametric_bootstrap(x.clean, mean, x.clean.center, x.clean.scale)
x.clean.bsp.meant <-  parametric_bootstrap(x.clean, mean, x.clean.center, x.clean.scale, trim=0.05)

data.frame(
  Sample = c(rep("x", 3), rep("x.clean", 3)),
  Estimator = c("Median", "Mean", "Mean Trimmed") %>% rep(2),
  Standard.Error = c(x.bs.median$se, x.bs.mean$se, x.bs.meant$se, x.clean.bs.median$se, x.clean.bs.mean$se, x.clean.bs.meant$se) %>% sapply(round, 4),
  Confidence.Intervals=list(x.bs.median$ci, x.bs.mean$ci, x.bs.meant$ci, x.clean.bs.median$ci, x.clean.bs.mean$ci, x.clean.bs.meant$ci) %>% sapply(function(c) paste(c[1] %>% round(4), ", ", c[2] %>% round(4))),
  Bias.Corrected.Estimate = c(x.bsp.median$bias_corrected_estimate, x.bsp.mean$bias_corrected_estimate, x.bsp.meant$bias_corrected_estimate, x.clean.bsp.median$bias_corrected_estimate, x.clean.bsp.mean$bias_corrected_estimate, x.clean.bsp.meant$bias_corrected_estimate)
) %>% 
  print() %>% 
  kable()



```




